\section{Limitations}
\label{sec:limitations}

CoMem is an efficiency and \emph{ultra-long-context} method, not an
accuracy state-of-the-art. Inside the backbone's extrapolation window
($\le$64k, where the full context still fits under the native positional
range), full-context recompute---our KV-Direct reference---matches or
exceeds CoMem: when the document fits, feeding all of it at full depth is
simply the better option. The clearest in-window gaps are on tasks that
reward exact localization over an intact context. On BABILong
(Table~\ref{tab:babilong}) single- and double-fact localization favors
KV-Direct: qa1 at 4k--32k is 75/80/72/62 for KV-Direct versus 66/63/57/28
for CoMem, and qa2 (the hardest double-fact setting) stays low for CoMem
across lengths. These are honest in-window losses; CoMem's advantage only
appears once the packed context exceeds the native window (\S\ref{sec:experiments}).

The watershed at which CoMem overtakes full context is backbone-dependent,
tracking roughly twice the native positional window. On Llama-3
\citep{llama3} (native 8k) full-context already collapses at 16k
($=$0) while CoMem holds; on Qwen3-8B \citep{qwen3} (native 40{,}960)
full-context survives to 64k and only collapses at 128k. A backbone with
strong length extrapolation therefore pushes the crossover out and narrows
the regime in which CoMem is the only viable choice; weaker-extrapolating
backbones benefit earlier and more.

Multi-hop reasoning over reference chains was a genuine weakness of
single-pass retrieval: on RULER variable-tracking a one-shot BM25 selector
scores only 27.6 at 16k and degrades further at 32k, because a single
retrieval pass cannot recover an entire chain of variable bindings. This
limitation is now \emph{largely retired} by the forward-free
\texttt{iter\_bm25} selector, which follows the reference chain and lifts
variable-tracking to 95.2/93.8/96.8 at 8k/16k/32k (Table~\ref{tab:itervt}),
non-degrading with length. Two caveats remain. First, iteration helps only
on chained multi-hop tasks and is scoped away from independent-needle tasks,
where it slightly hurts (multikey 92 vs.\ 97--99 single-pass) or is a no-op
(single needle 100$=$100). Second, the \emph{decode-time cost} of iterative
retrieval---each hop issues an additional retrieval and read---is not yet
characterized; the reported gains are accuracy-only.

Decode is slow. Because every generated token recomputes layers[$j{:}L$]
with full attention over the entire read pack, per-step decode latency is
$\sim$2.4\,s at chunk size 512 and rises to $\sim$5.5\,s at chunk size 1024,
against $\sim$0.3--0.5\,s for full attention (Table~\ref{tab:eff}). CoMem's
wins are in prefill and memory (constant $\sim$18\,GB, 7.83$\times$ prefill
speedup at 128k), not in per-token decode. Caching the resumed upper-band KV
across decode steps would remove most of this cost and is left to future work.

We are careful about what ``compressibility'' means here. CoMem's storage
saving is on the \emph{layer/feature axis}: we cache a single depth-$j$
hidden per chunk---roughly $1/(2L)$ of the full-depth KV---rather than
retaining all layers. We do \emph{not} claim that shallower hidden states
are inherently more compressible along the value axis (e.g.\ under int4
quantization); that stronger hypothesis was in fact falsified, as
int4-quantizing the shallow $j{=}6$ hidden yields a relative error of 0.84.
The gain comes from storing one layer, not from shallow states quantizing
better.

CoMem inherits the backbone's behavior on adversarial refusal. On LoCoMo
(Table~\ref{tab:locomo}), the adversarial category cat5---which requires the
model to refuse unanswerable questions---is weak (F1/acc $\approx$1.6),
mirroring the stock backbone's reluctance to abstain rather than any
retrieval failure. Handling such queries would require an explicit refusal
mechanism, which CoMem does not add.

Finally, some evaluation settings are calibrated for cost rather than uniform
sample counts, and we flag them for honesty. The RULER head-to-head
(Table~\ref{tab:h2h}) is an $n{=}50$ cut, whereas most other results use
$n{=}100$; the topk$\times$length grid is sparse (only 32k carries a complete
topk curve); and we report three distinct 128k niah\_multikey values---98
(best-topk, tk8), 96 (fixed tk12, head-to-head), and 84 (length-adaptive
tk24, scaling)---which differ by topk convention, not by contradiction, each
annotated with its topk setting in the relevant table.
